Separating random and deterministic sources of computational noise in explore-exploit decisions

### 2024-05-16 manuscript/2024 - Plot Comp Bio/manuscript_DeterministicRandomNoise.tex
u

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 14$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 86$\%$ of the variability is truly `random,' although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}
	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

@@

	\section*{Results}	
	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of the `Horizon Task' \citep{wilson2014} to show the influence of stimulus-driven `deterministic noise' vs non-stimulus-driven `random noise' in explore-exploit decisions (Figure \ref{fig:taskfig}). In this task, participants make a series of choices between two slot machines, or `one-armed bandits,' that pay out probabilistic rewards. They are asked to choose between the two bandits to maximize the total rewards. One bandit always has a higher mean payout than the other. Participants need to try each bandit a few times to learn about the distribution of payout from that bandit. Because they are initially unsure as to the mean payoff of each bandit, this task requires that participants carefully balance exploration of the lesser known bandit with exploitation of the better known bandit to maximize their overall rewards.

@@

\subsubsection*{Model validation}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters. Full details of this analysis is presented in the Supplementary Materials section 2.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\textwidth]{figures/RDBayes_parameterrecovery_subject_examplesession_randet.jpg}
		\caption{Parameter recovery analysis for random (A,B) and deterministic (C,D) noises in the two horizons.}
		\label{fig:paramrecover_main}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior.jpg}
	\caption{Model based analysis showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and increase with horizon (B, D).  }
	%BOB SAYS: However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.
	\label{fig:mb1}
\end{center}
\end{figure}

@@

\section*{Discussion}
In this paper, we investigated whether random exploration is really random or whether it is driven deterministically by aspects of the stimulus we have previously ignored when measuring `decision noise.'  Using a version of the Horizon Task with repeated games, we found evidence that at least some of the noise in random exploration could be explained by such `deterministic noise.' In particular, we found that deterministic noise accounted for around 14\% of the overall variability in people's behavior.

>>>>

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 14$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 86$\%$ of the variability is truly `random', although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}
	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

@@

	\section*{Results}	
	\subsection*{The Repeated-Games Horizon Task}
	We used a modified version of the `Horizon Task' \citep{wilson2014} to show the influence of stimulus-driven `deterministic noise' vs non-stimulus-driven `random noise' in explore-exploit decisions (Figure \ref{fig:taskfig}). In this task, participants make a series of choices between two slot machines, or `one-armed bandits', that pay out probabilistic rewards. They are asked to choose between the two bandits to maximize the total rewards. One bandit always has a higher mean payout than the other. Participants need to try each bandit a few times to learn about the distribution of payout from that bandit. Because they are initially unsure as to the mean payoff of each bandit, this task requires that participants carefully balance exploration of the lesser known bandit with exploitation of the better known bandit to maximize their overall rewards.

@@

\subsubsection*{Model validation}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters. Full details are presented in the Supplementary Materials section 2.

@@

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=.95\textwidth]{figures/RDBayes_parameterrecovery_subject_examplesession_randet.jpg}
		\caption{Parameter recovery analysis for random (A,B) and deterministic (C,D) noises in the two horizons.}
		\label{fig:paramrecover_main}
	\end{center}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
	\includegraphics[width=\textwidth]{figures/RDBayes_hyperprior.jpg}
	\caption{Model based analysis showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and increase with horizon (B, D).  }
	%BOB SAYS: However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.
	\label{fig:mb1}
\end{center}
\end{figure}

@@

\section*{Discussion}
In this paper, we investigated whether random exploration is really random or whether it is driven deterministically by aspects of the stimulus we have previously ignored when measuring `decision noise'.  Using a version of the Horizon Task with repeated games, we found evidence that at least some of the noise in random exploration could be explained by such `deterministic noise'. In particular, we found that deterministic noise accounted for around 14\% of the overall variability in people's behavior.
